
Trellis Graphs
Trellis graphs are avialable throught the lattice package. A trellis graph displays a variable or the relationship between variables, conditioned on one or more other variables. Trellis graphs are available for a wide variety of plot types.
The typical format is
graph_type(formula, data=)
where graph_type is selected from the listed below. formula specifies the variable(s) to display and any conditioning variables . For example ~x|A means display numeric variable x for each level of factor A. y~x | A*B means display the relationship between numeric variables y and x separately for every combination of factor A and B levels.~x means display numeric variable x alone.
graph_type
description
formula examples
barchart
bar chart
x~A or A~x
bwplot
boxplot
x~A or A~x
cloud
3D scatterplot
z~x*y|A
contourplot
3D contour plot
z~x*y
densityplot
kernal density plot
~x|A*B
dotplot
dotplot
~x|A
histogram
histogram
~x
levelplot
3D level plot
z~y*x
parallel
parallel coordinates plot
dataframe
splom
scatterplot matrix
dataframe
stripplot
strip plots
A~x or x~A
xyplot
scatterplot
y~x|A
wireframe
3D wireframe graph
z~y*x
Here are some examples. They use the car data (mileage, weight, number of gears, number of cylinders, etc.) from the mtcars dataframe.
# Trellis Examples 
library(lattice) 
attach(mtcars)

# create factors with value labels 
gear.f<-factor(gear,levels=c(3,4,5),
labels=c("3gears","4gears","5gears")) 
cyl.f <-factor(cyl,levels=c(4,6,8),
labels=c("4cyl","6cyl","8cyl")) 

# kernel density plot 
densityplot(~mpg, 
main="Density Plot", 
xlab="Miles per Gallon")

# kernel density plots by factor level 
densityplot(~mpg|cyl.f, 
main="Density Plot by Number of Cylinders",
xlab="Miles per Gallon")

# kernel density plots by factor level (alternate layout) 
densityplot(~mpg|cyl.f, 
main="Density Plot by Numer of Cylinders",
xlab="Miles per Gallon", 
layout=c(1,3))

# boxplots for each combination of two factors 
bwplot(cyl.f~mpg|gear.f,
ylab="Cylinders", xlab="Miles per Gallon", 
main="Mileage by Cylinders and Gears", 
layout=(c(1,3))

# scatterplots for each combination of two factors 
xyplot(mpg~wt|cyl.f*gear.f, 
main="Scatterplots by Cylinders and Gears", 
ylab="Miles per Gallon", xlab="Car Weight")

# 3d scatterplot by factor level 
cloud(mpg~wt*qsec|cyl.f, 
main="3D Scatterplot by Cylinders") 

# dotplot for each combination of two factors 
dotplot(cyl.f~mpg|gear.f, 
main="Dotplot Plot by Number of Gears and Cylinders",
xlab="Miles Per Gallon")

# scatterplot matrix 
splom(mtcars[c(1,3,4,5,6)], 
main="MTCARS Data")


click to view
Note, as in graph 1, that you specifying a conditioning variable is optional. The difference between graphs 2 & 3 is the use of the layout option to contol the placement of panels.

Customizing Trellis Graphics


Unlike other R graphs, the trellis graphs described here are not effected by many of the options set in the par( ) function. To view the options that can be changed, look athelp(xyplot). It is frequently easiest to set these options within the high level plotting functions described above. Additionally, you can write functions that modify the rendering of panels. Here is an example.
# Customized Trellis Example
library(lattice)
panel.smoother <- function(x, y) {
	panel.xyplot(x, y) # show points 
	panel.loess(x, y)  # show smoothed line 
}
attach(mtcars)
hp <- cut(hp,3) # divide horse power into three bands 
xyplot(mpg~wt|hp, scales=list(cex=.8, col="red"),
panel=panel.smoother,
xlab="Weight", ylab="Miles per Gallon", 
main="MGP vs Weight by Horse Power")

