\newpage
\section{Probability Distribution used for Extreme Value Theory}
\begin{itemize}
\item The GEV distribution
\item The Pareto distribution
\end{itemize}
%----------------------------------------------------------------------------------------%
\subsection{The Generalized Extreme Value (GEV) Distribution}
The generalized extreme value (GEV) distribution is a family of continuous probability distributions developed within extreme value theory to combine the Gumbel, Fréchet and Weibull families also known as type I, II and III extreme value distributions.

By the extreme value theorem the GEV distribution is the limit distribution of properly normalized maxima of a sequence of independent and identically distributed random variables. Because of this, the GEV distribution is used as an approximation to model the maxima of long (finite) sequences of random variables.

In some fields of application the generalized extreme value distribution is known as the Fisher–Tippett distribution, named after R. A. Fisher and L. H. C. Tippett. However usage of this name is sometimes restricted to mean the special case of the Gumbel distribution.
%----------------------------------------------------------------------------------------%
\subsection{The Pareto Distribution}

In hydrology the Pareto distribution is applied to extreme events such as annually maximum one-day rainfalls and river discharges. The blue picture illustrates an example of fitting the Pareto distribution to ranked annually maximum one-day rainfalls showing also the 90\% confidence belt based on the binomial distribution. The rainfall data are represented by plotting positions as part of the  \textbf{\emph{cumulative frequency analysis}}.

\subsection{The Gumbel Distribution}
The extreme value type I distribution has two forms. One is based
on the smallest extreme and the other is based on the largest
extreme. We call these the minimum and maximum cases,
respectively. Formulas and plots for both cases are given. The
extreme value type I distribution is also referred to as the
Gumbel distribution.

%----------------------------------------------------------------------------------------%